Practical Optional Types for Clojure

### 2015-02-26 formal-model.tex
section3 intro

The first insight about occurrence typing is that
propositions can be used to capture type information about our programs,
in particular by referencing components or subcomponents of the runtime environment
and relating them to types.
\emph{Type Propositions} make assertions like ``variable \clj{x} is of type \clj{Number}'' or
``variable \clj{x} is not \clj{nil}''---
in our logical system we write these as
{\isprop{\Number}{\x{}}}
and {\notprop{\ValueNil}{\x{}}}. 
The other propositions are standard: the implication, conjunction,
disjunction, trivial (\topprop{}) and impossible (\botprop{}) propositions
(figure~\ref{main:figure:termsyntax}).

@@

The second insight is that we can replace the traditional 
representation of a
type environment of a map from variables to types
with a sequence of propositions, written \propenv{}. 
Instead of mapping \clj{x} to
the type \clj{Number}, we instead use the proposition {\isprop{\Number}{\x{}}}.

The third insight is that we
can use a logic system to learn new information about our programs.
The key component of the logic system L-Update uses type propositions
to derive new propositions.
$$
  {\LUpdate}
$$
For example, with L-Update we can use the knowledge of
\inpropenv{\propenv{}}{\isprop{\Union{\ValueNil{}}{\Number{}}}{\x{}}}
and 
\inpropenv{\propenv{}}{\notprop{\ValueNil{}}{\x{}}}
to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
We cover \updateliteral{} in section~\ref{sec:updateformalsection}.

The fourth insight is a consequence of
programmers in dynamic languages
using implicit type-based reasoning based on the result of
conditional tests.
For instance, example~\ref{example:conditionalflow}
only utilises \clj{x} once
the programmer is convinced it is safe to do so based whether
\clj{(number? x)}
is 
a true or false value. The takeaway is that every expression has
two relevant pieces of information to the logic system: a `then' proposition
for when it reduces to a true value, and an `else' proposition
when it reduces to a false value---for \clj{(number? x)}
they are {\isprop{\Number}{\x{}}} and {\notprop{\Number}{\x{}}}
respectively.
%\ref{main:figure:typingrules}

@@

Figure~\ref{main:figure:typingrules} contains the standard typing rules.
We can now demonstrate how to check example~\ref{example:conditionalflow}.
$$
\judgement{\isprop{\clj{(U nil Num)}}{\x{}}}{\hastype{\clj{(number? x)}}{\clj{Boolean}}}{\filterset{\isprop{\clj{Num}}{\x{}}}{\notprop{\clj{Num}}{\x{}}}}{\emptyobject{}}
$$
$$
\judgement{\isprop{\clj{Num}}{\x{}}}{\hastype{\clj{(inc x)}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$
$$
\judgement{\isprop{\clj{Num}}{\x{}}}{\hastype{\clj{x}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\x{}}
$$
$$
\judgement{\isprop{\clj{Num}}{\x{}}}{\hastype{\clj{0}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$
By T-Abs

>>>>

The first insight about occurrence typing is that
propositions can be used to capture type information about our programs,
in particular by referencing components or subcomponents of the runtime environment
and relating them to types.
\emph{Type Propositions} make assertions like ``variable \clj{x} is of type \clj{Number}'' or
``variable \clj{x} is not \clj{nil}''---
in our logical system we write these as
{\isprop{\clj{Number}}{\x{}}}
and {\notprop{\clj{nil}}{\x{}}}. 
The other propositions are standard: the implication, conjunction,
disjunction, trivial (\topprop{}) and impossible (\botprop{}) propositions
(figure~\ref{main:figure:termsyntax}).

@@

The second insight is that we can replace the traditional 
representation of a
type environment of a map from variables to types
with a sequence of propositions, written \propenv{}. 
Instead of mapping \clj{x} to
the type \clj{Number}, we instead use the proposition {\isprop{\clj{Number}}{\x{}}}.

The third insight is that we
can use a logic system to learn new information about our programs.
The key component of the logic system L-Update uses type propositions
to derive new propositions.
$$
  {\LUpdate}
$$
For example, with L-Update we can use the knowledge of
\inpropenv{\propenv{}}{\isprop{\clj{(U nil Number)}}{\x{}}}
and 
\inpropenv{\propenv{}}{\notprop{\clj{nil}}{\x{}}}
to derive \inpropenv{\propenv{}}{\isprop{\clj{Number}}{\x{}}}.
(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
We cover \updateliteral{} in section~\ref{sec:updateformalsection}.

The fourth insight is a consequence of
programmers in dynamic languages
using implicit type-based reasoning based on the result of
conditional tests.
For instance, example~\ref{example:conditionalflow}
only utilises \clj{x} once
the programmer is convinced it is safe to do so based whether
\clj{(number? x)}
is 
a true or false value. The takeaway is that every expression has
two extra pieces of information relevant to the logic system: a `then' proposition
for when it reduces to a true value, and an `else' proposition
when it reduces to a false value---for \clj{(number? x)}
the then proposition is {\isprop{\clj{Number}}{\x{}}} and 
the else proposition is {\notprop{\clj{Number}}{\x{}}}.
%\ref{main:figure:typingrules}

@@

Figure~\ref{main:figure:typingrules} contains the standard typing rules.
We can now demonstrate how to check example~\ref{example:conditionalflow}.
We abbreviate \clj{Num} as \clj{N},  \clj{Boolean} as \clj{B} and \clj{number?} as \clj{n?}.
The test is checked with T-App, 
$$
\judgement{\isprop{\clj{(U nil N)}}{\x{}}}{\hastype{\clj{(n? x)}}{\clj{B}}}{\filterset{\isprop{\clj{N}}{\x{}}}{\notprop{\clj{N}}{\x{}}}}{\emptyobject{}}
$$
\clj{n?} has type
$$
{\ArrowOne{\x{}}{\clj{Any}}{\clj{B}}
        {\filterset{\isprop{\clj{N}}{\x{}}}{\notprop{\clj{N}}{\x{}}}}{\emptyobject{}}}
$$
By T-Local.
$$
\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{x}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\x{}}
$$
Now we check both branches.
$$
\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{(inc x)}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$
$$
\judgement{\notprop{\clj{N}}{\x{}}}{\hastype{\clj{0}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$
We can now check the conditional with T-If.
$$
\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{(if (n? x) (inc x) 0)}}{\clj{N}}}{\filterset{\orprop{\isprop{\clj{N}}{\x{}}}{\topprop{}}}{\orprop{\notprop{\clj{N}}{\x{}}}{\topprop{}}}}{\emptyobject{}}
$$
Finally the function can be checked with T-Abs.
$$
\judgement{}{\hastype{\clj{(fn ...)}}
                                             {\ArrowOne{\x{}}{\clj{(U nil N)}}{\clj{N}}
        {\filterset{\orprop{\isprop{\clj{N}}{\x{}}}{\topprop{}}}{\orprop{\notprop{\clj{N}}{\x{}}}{\topprop{}}}}{\emptyobject{}}}}
  {\filterset{\topprop{}}{\botprop{}}}{\emptyobject{}}
$$
By T-Abs
